Ensemble Stability of Biological Networks

### 2014-09-18 paper.tex
integrate changes from sharelatex

For the case of continuous deterministic observables, the above process can be made more precise by associating a manifold $M_i$ to each observable, a directed graph $G$ to the interdependencies, and a vector field $V \in T(\prod_i M_i)$ over the space determined by taking the product of the manifolds associated to the collection of observables, satisfying these interdependencies for each observable  \cite{Deville}. For example, if we have two observables $\{x_1,x_2\}$ where the domains in which they are quantified are given by manifolds $\{M_1,M_2\}$ such that $x_1 \in M_1 \equiv \mathbb{R}^1$ and $x_2 \in M_2 \equiv \mathbb{R}^1$, a directed graph $G_X$ describing the dependencies between these observables and vector field $V$ with components $\{F_1,F_2\}$ defined on $M_1 \otimes M_2 \equiv \mathbb{R}^2$ satisfying the dependencies determined by $G_X$. If the system under consideration has the graph given in \ref{fig:examplesystemmodules}A
% \begin{center}
% \input{tex/examplehamdyngraph}
% \end{center}
with adjacency matrix
$$
\adj(G_X) = \begin{pmatrix}
0 & 1 \\
1 & 1
\end{pmatrix}
$$
then for the system $X = \{G_X, M_X, F_X\}$, where $M_X \equiv \{M_1,M_2\}$ and $F_X \equiv \{F_1,F_2\}$ such that $F_1$ is constrained to be a function of $x_2$ and $F_2$ is a function of both $x_1$ and $x_2$ yielding the flow equations
\begin{align*}
\frac{dx_1}{dt} & = F_1(x_2)\\
\frac{dx_2}{dt} & = F_2(x_1,x_2)
\end{align*}

@@

Another question one can ask is that of robustness, which measures the
probability that, if some property holds for a set of parameters, it will
continue to hold if we make a random perturbation about those parameter values.
To define robustness, we will introduce, in addition to the probability
distribution $\mu$ described above, a family of probability distributions
$\mu' \colon S \times \mathcal{B}(S) \to [0,1]$.  Then, given a binary random
variable $q$, we we define its robustness as the following conditional proability:
\[
  \mathrm{Robust} (q) =
  \frac{\int_S d\mu(x) \int_S d\mu'(y) \mathbf{1}_q(x) \mathbf{1}_q(y)}
  {\int_S d\mu(x) \mathbf{1}_q(x)}
\]

@@

\section{Stability criteria}
Suppose that $M$ is an $n \times n$ matrix with real coefficients.
Motivated by the discussion of \ref{sec:stabanalbn}, we will call
the matrix $M$ stable if all its eigenvalues have negative real part.
Define the characteristic polynomial as $\chi(M)(z) = \det(zI - M)$.
Denote the coefficients of $\chi(M)$ as $s_j$ and its roots as $r_j$,
writing $p(x) = x^n + \sum_{k=1}^n s_k x^{n-k} = \prod_{k=1}^n (x-r_k)$.
Then, since the roots of $\chi$ are the eigenvalues of $M$, asking that
$M$ be stable is equivalent to asking that the roots of $\chi$ all have
negative real part.  We will formulate conditions for this to happen in
terms of the coefficients of $\chi$.

@@

\section{Paths and eigenvalues}
Next, we will want to relate prpoperties of our matrix and its eigenvalues
to  properties of the associated graph $G_A$.  To do this, we will first
examine the category of paths on $G_A$, and derive a formula expressing
the chararacteristic polynomial as a sum over cycles.

@@

Next, we introduce two basic functors from $\Path (G)$.  One of these is
$\length \colon \Path(G) \to \mathbb{N}$ (regarding the natural numbers
as a category with a single object) which sends each path to its length.
The other functor is the weight product functor.  Given an assignement
$w \colon \Path(G) \to \mathbb(R)$ of weights to the edges of our digraph,
define $\wprod \colon \Path(G) \to \mathbb{R}$ to be the function which sends
each path to the product of the weights associated to the edges which
that path traverses.  Given the way that graphs are associated to matrices,
weights on edges correspond to matrix entries and, given a path
$v_1, v_2, v_3 \ldots$, we have $\wprod(v_1, v_2, v_3 \ldots) =
m_{v_1 v_2} m_{v_2 v_3} \cdots$.  Going in the opposite direction, we
note that any functor from $\Path(G)$ to $\mathbb{R}$ may be regarded as
a weighted product for a suitable choice of weights.  We extend $\wp$ to
$\mathcal{A}(G)$ by linearity; the result is an algebra homomorphism.

@@

As an illustration of this formula, we will use it to compute the characteristic
polynomial of the matrix associated to the labelled graph in figure ??.  We begin
by noting that the closed paths of this graph can be described by the following
regular expressions:
\begin{align*}
\Path (1,1) &: (e + bd + bca)^* \\
\Path (2,2) &: (ae^*(be^*d)^*bc)^* \\
\Path (3,3) &: ((ca + d)e^*b)^* \\
\end{align*}
Since these regular expressions are non-redundant, we can obtain generating
functions for the paths by considering $a,b,c,d,e$ as non-commutative
variables and replacing the Kleene star $x^*$ by $(1 - x)^{-1}$:
\begin{align*}
\Path (1,1) &: (1 - e - bd - bca)^{-1} \\
\Path (2,2) &: (1 - a(1-e)^{-1}(1 - b(1-e)^{-1}d)^{-1}bc)^{-1} \\
\Path (3,3) &: (1 - (ca + d)(1 - e)^{-1}b)^{-1} \\
\end{align*}
To obtain the characteristic polynomial, we make the substitutions
$a \mapsto m_{12}/z, b \mapsto m_{31}/z, c \mapsto m_{23}/z, d \mapsto m_{13}/z,
e \mapsto m_{11}/z$ and simplify, using the fact that the matrix entries
commute:
\begin{align*}
\Path (1,1) &: \frac{z^3}
{z^3 - m_{11} z^2 - m_{13} m_{31} z - m_{12} m_{23} m_{31} } \\
\Path (2,2) &: \frac{z(z^2 - m_{11} z - m_{13} m_{31})}
{z^3 - m_{11} z^2 - m_{13} m_{31} z - m_{12} m_{23} m_{31} }  \\
\Path (3,3) &: \frac{z^2(z^2 - m_{11})}
{z^3 - m_{11} z^2 - m_{13} m_{31} z - m_{12} m_{23} m_{31} } \\
\end{align*}
Adding these three quantitites together, we obtain
\[
\frac{z(3 z^2 - 2 m_{11} z - m_{13} m_{31})}
{z^3 - m_{11} z^2 - m_{13} m_{31} z - m_{12} m_{23} m_{31} },
\]
which we recognize as $\chi'(z)/\chi(z)$, where
$\chi(z) = z^3−m_{11} z^2 - m_{13} m_{31} z - m_{12} m_{23} m_{31}$.
finally, we note that this polynomial is indeed the characteristic
polynomial of the matrix
\[
 \begin{matrix}
   m_{11} & m_{12} & m_{13} \\
   0      & 0      & m_{23} \\
   m_{31} & 0      & 0 \\
 \end{matrix}
\]
corresponding to our digraph.

>>>>

For the case of continuous deterministic observables, the above process can be made more precise by associating a manifold $M_i$ to each observable, a directed graph $G$ to the interdependencies, and a vector field $V \in \Gamma T(\prod_i M_i)$ over the space determined by taking the product of the manifolds associated to the collection of observables, satisfying these interdependencies for each observable  \cite{Deville}. For example, if we have two observables $\{x_1,x_2\}$ where the domains in which they are quantified are given by manifolds $\{M_1,M_2\}$ such that $x_1 \in M_1 \equiv \mathbb{R}^1$ and $x_2 \in M_2 \equiv \mathbb{R}^1$, a directed graph $G_X$ describing the dependencies between these observables and vector field $V$ with components $\{F_1,F_2\}$ defined on $M_1 \otimes M_2 \equiv \mathbb{R}^2$ satisfying the dependencies determined by $G_X$. If the system under consideration has the graph given in \ref{fig:examplesystemmodules}A
% \begin{center}
% \input{tex/examplehamdyngraph}
% \end{center}
with adjacency matrix
$$
\adj(G_X) = \begin{pmatrix}
0 & 1 \\
1 & 1
\end{pmatrix}
$$
then for the system $X = \{G_X, M_X, F_X\}$, where $M_X \equiv \{M_1,M_2\}$ and $F_X \equiv \{F_1,F_2\}$ such that $F_1$ is constrained to be a function of $x_2$ and $F_2$ is a function of both $x_1$ and $x_2$ yielding the flow equations
\begin{align*}
\frac{dx_1}{dt} & = F_1(x_2)\\
\frac{dx_2}{dt} & = F_2(x_1,x_2)
\end{align*}

@@

Another question one can ask is that of robustness, which measures the
probability that, if some property holds for a set of parameters, it will
continue to hold if we make a random perturbation about those parameter values.
To define robustness, we will introduce, in addition to the probability
distribution $\mu$ described above, a family of probability distributions
$\mu' \colon S \times \mathcal{B}(S) \to [0,1]$.  Then, given a binary random
variable $q$, we we define its robustness as the following conditional proability:
\[
  \mathrm{Robust} (q) =
  \frac{\int_S d\mu(x) \int_S d\mu'(x,y) \mathbf{1}_q(x) \mathbf{1}_q(y)}
  {\int_S d\mu(x) \mathbf{1}_q(x)}
\]

@@

\section{Component decomposition and robustness}

The decomposition of a digraph into strongly connected components 
corresponds to a block triangular decomposition of matrices
describle by that digraph.  Say that the graph $G$ has strongly
connected components $C_1, C_2, \ldots C_n$, which have been
labelled in such a way that there are no links from vertices in
component $C_i$ to component $C_j$ when $i < j$.  Label the
vertices in such a way that $v_1, v_{n_1}$ belong to $C_1$,
$v_1, v_{n_1}$ belong to $C_2$, etc.  Then if $M$ is a matrix
consistent with $G$, it must be block triangular.  [explain better]

An important consequence of this is that edges which go between
different strongly connected components do not contribute to
determining stability.  Hence, if we resample a matrix entry
corresponding to such an edge, the conditional probability 
of the matrix remaining stable upon resampling is 1.  Hence, if
our graph $G$ has strongly connected components $C_1, C_2, \ldots C_n$
where $C_i$ has $v_i$ vertices and connectivity $c_1$ and there
are $\ell_{ij}$ links between components $i$ and $j$, then the
total robustness to resampling a random entry is
\[
R(G) = \frac{\sum_{i=1}^n \sum_{j=1}^n \ell_{ij} + \sum_{i=1}^n c_i R_i}
            {\sum_{i=1}^n \sum_{j=1}^n \ell_{ij} + \sum_{i=1}^n c_i}
\]

\[
R_{\mathrm{max}} = \frac{\frac{1}{2} ((\sum_i n_i)^2 - \sum_i n_i^2) + \sum_i c_i R_i}
            {\frac{1}{2} ((\sum_i n_i)^2 - \sum_i n_i^2) + \sum_i c_i}
\]

\section{Stability criteria}
\todo{Since there is no point in publishing proofs of results contemporary
with Darwin's Origin of Species, this section will be replaced with a
reference to Gantmacher.}
Suppose that $M$ is an $n \times n$ matrix with real coefficients.
As described in \ref{sec:stabanalbn}, we call
the matrix $M$ stable if all its eigenvalues have negative real part.
Define the characteristic polynomial as $\chi(M)(z) = \det(zI - M)$.
Denote the coefficients of $\chi(M)$ as $s_j$ and its roots as $r_j$,
writing $p(x) = x^n + \sum_{k=1}^n s_k x^{n-k} = \prod_{k=1}^n (x-r_k)$.
Then, since the roots of $\chi$ are the eigenvalues of $M$, asking that
$M$ be stable is equivalent to asking that the roots of $\chi$ all have
negative real part.  Hence, let us call a polynomial stable if all its
roots have negative real part.  We will formulate conditions for this to 
happen in terms of the coefficients of $\chi$.

@@

\begin{lemma} 
If $n < 6$, then a polynomial $p$ is stable if and only
if $\Delta > 0$ and $s_1 > 0$ for all $i$.
\end{lemma}

\begin{proof}
Suppose that, to the contrary, there exists a polynomial which satisfies
the inequalities $\Delta > 0$ and $s_1 > 0$ but which is not stable.
Consider a straight line connecting this polynomial to a stable polynomial,
say $(x - 1)^n$.

Suppose that $p$ is a polynomial for which $\Delta = 0$ and $s_i > 0$ but
which is not on the boundary of the stable region.  In order to have 
$\Delta = 0$, we must either have a pair of real roots which sum to zero 
or a pair of complex conjugate roots with zero real part.  In the former
case, we would have $x^2 - r^2$ as a factor of our polynomial and in the
latter, we would have $x^2 + r^2$ as a factor.

Suppose that $x^2 - r^2$ is a factor.  Then we have
\[
p = (x^2 - r^2) \left(x^{n-2} + \sum_{k=1}^{n-3} c_k x^{n-k-2} \right) =
x^n + \sum_{k=3}^{n-3} (c_k - r^2 c_{k-2}) x^{n-k} + 
\]

Suppose that $x^2 + r^2$ is a factor. Then we have one of the following
when $2 < n < 6$
\begin{align*}
(x^2 + r^2)(x + a_1) &=
x^3 + a_1 x^3 + r^2 x^2 + a_1 r^2 \\
(x^2 + r^2)(x^2 + a_1 x + a_2) &=
x^4 + a_1 x^3 + (a_2 + r^2) x^2 + a_1 r^2 x + a_2 r^2 \\
(x^2 + r^2)(x^3 + a_1 x^2 + a_2 x + a_3) &=
x^5 + a_1 x^4 + (a_2 + r^2) x^3 + (a_3 + a_1 r^2) x^2 + a_2 r^2 x + a_3 r^2
\end{align*}
Inspecting these, we see that, if the coefficients of $p$ are positive,
then $a_i > 0$ for all $i$.

$\ldots$
\end{proof}

For $n = 6$, this argument breaks down because, for instance, we have
the polynomial $x^6 + x^5 +\frac{1}{2} x^4 + 2x^3 + \frac{1}{2} x^2 +
x + 1$  which factors as  $(x^2 + 1)(x^4 + x^3 - \frac{1}{2} x^2 + x + 1)$.
This polynomial has positive cofficients and $\Delta = 0$ but does not
lie on the boundary of the stable region because the second factor has
a negative coefficient, hence has a root with positive real part.  Thus,
for $n = 6$ and higher, we need to supplement our conditions with
additional inequalities (which may be taken to be linear) to exclude
such possibilitiies.  However, we will not pursue this topic further
in this work because we will only be making computations with matrices
of size less than $6$.

\section{Paths and eigenvalues}
\todo{I am thinking that we might want to jettison this section from
the published version and leave it as private notes for future work.
As a substitute for the published version, I made a section in front 
of it which takes a more direct route to proving what is needed for
the purposes of showing that the characteristic polynomial factors and
making our arguments about total orders being the most stable.  We
should discuss this point tomorrow.}

Next, we will want to relate prpoperties of our matrix and its eigenvalues
to  properties of the associated graph $G_A$.  To do this, we will first
examine the category of paths on $G_A$, and derive a formula expressing
the chararacteristic polynomial as a sum over cycles.

@@

Next, we introduce two basic functors from $\Path (G)$.  One of these is
$\length \colon \Path(G) \to \mathbb{N}$ (regarding the natural numbers
as a category with a single object) which sends each path to its length.
The other functor is the weight product functor.  Given an assignement
$w \colon \Path(G) \to \mathbb(R)$ of weights to the edges of our digraph,
define $\wprod \colon \Path(G) \to \mathbb{R}$ to be the function which sends
each path to the product of the weights associated to the edges which
that path traverses.  Given the way that graphs are associated to matrices,
weights on edges correspond to matrix entries and, given a path
$v_1, v_2, v_3 \ldots$, we have $\wprod(v_1, v_2, v_3 \ldots) =
m_{v_1 v_2} m_{v_2 v_3} \cdots$.  Going in the opposite direction, we
note that any functor from $\Path(G)$ to $\mathbb{R}$ may be regarded as
a weighted product for a suitable choice of weights.  We extend $\wprod$ to
$\mathcal{A}(G)$ by linearity; the result is an algebra homomorphism.

@@

As an illustration of this formula, we will use it to compute the characteristic
polynomial of the matrix associated to the labelled graph in figure ??.  We begin
by noting that the closed paths of this graph can be described by the following
regular expressions:
\begin{align*}
\Path (1,1) &: (e + bd + bca)^* \\
\Path (2,2) &: (ae^*(be^*d)^*bc)^* \\
\Path (3,3) &: ((ca + d)e^*b)^* \\
\end{align*}
Since these regular expressions are non-redundant, we can obtain generating
functions for the paths by considering $a,b,c,d,e$ as non-commutative
variables and replacing the Kleene star $x^*$ by $(1 - x)^{-1}$:
\begin{align*}
\Path (1,1) &: (1 - e - bd - bca)^{-1} \\
\Path (2,2) &: (1 - a(1-e)^{-1}(1 - b(1-e)^{-1}d)^{-1}bc)^{-1} \\
\Path (3,3) &: (1 - (ca + d)(1 - e)^{-1}b)^{-1} \\
\end{align*}
To obtain the characteristic polynomial, we make the substitutions
$a \mapsto m_{12}/z, b \mapsto m_{31}/z, c \mapsto m_{23}/z, d \mapsto m_{13}/z,
e \mapsto m_{11}/z$ and simplify, using the fact that the matrix entries
commute:
\begin{align*}
\Path (1,1) &: \frac{z^3}
{z^3 - m_{11} z^2 - m_{13} m_{31} z - m_{12} m_{23} m_{31} } \\
\Path (2,2) &: \frac{z(z^2 - m_{11} z - m_{13} m_{31})}
{z^3 - m_{11} z^2 - m_{13} m_{31} z - m_{12} m_{23} m_{31} }  \\
\Path (3,3) &: \frac{z^2(z^2 - m_{11})}
{z^3 - m_{11} z^2 - m_{13} m_{31} z - m_{12} m_{23} m_{31} } \\
\end{align*}
Adding these three quantitites together, we obtain
\[
\frac{z(3 z^2 - 2 m_{11} z - m_{13} m_{31})}
{z^3 - m_{11} z^2 - m_{13} m_{31} z - m_{12} m_{23} m_{31} },
\]
which we recognize as $\chi'(z)/\chi(z)$, where
$\chi(z) = z^3−m_{11} z^2 - m_{13} m_{31} z - m_{12} m_{23} m_{31}$.
finally, we note that this polynomial is indeed the characteristic
polynomial of the matrix
\[
 \begin{pmatrix}
   m_{11} & m_{12} & m_{13} \\
   0      & 0      & m_{23} \\
   m_{31} & 0      & 0 \\
 \end{pmatrix}
\]
corresponding to our digraph.
